---
title: "Close Simulations"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read in simulation results

- 50 SliM simulations, each from one combination of parameters chosen through abc based on number of patches

- For each SLiM simulation:
  - 5 recapitation replicates
  - For each recapitation replicate:
    - 5 mutation replicates, with mutation rate drawn from uniform 2.0e-9, 5.6e-9
    - For each mutation replicate:
      - 5 replicates of choosing patches to match

```{r data, echo = FALSE, cache = TRUE}
library(tidyverse)
source("../data/helpers.R")
basedir <- "post_21000_2022-12-05"
# Observed heterozygosity and pairwise stats
observed_df <- read.csv(file.path(basedir, "cleaned_results/observed.csv"))
observed <- observed_df$x
names(observed) <- rownames(observed_df)
# Simulated heterozygosity and pairwise stats
stats <- read.csv(file.path(basedir,"cleaned_results/stats.csv"))
stats <- dplyr::select(stats, !X)
# Parameters for each simulation and replicate
rep_info <- read.csv(file.path(basedir,"cleaned_results/rep_info.csv"))
rep_info <- rep_info |> mutate(across(c(sim, recap, mut, sample), factor))
# Sampling locations
sample_locs <- read.csv(file.path(basedir,"cleaned_results/sample_locs.csv"))
# Stats for all simulations
all_stats <- read.csv(file.path(basedir,"cleaned_results/all_stats.csv"))
# Pairwise stats for all simulations
all_pairstats <- read.csv(file.path(basedir,"cleaned_results/all_pairstats.csv"))

observed_stats <- read.csv(file.path(basedir,"cleaned_results/observed_stats.csv"))
observed_pairstats <- read.csv(file.path(basedir,"cleaned_results/observed_pairstats.csv"))

```

## Plot a subset of 10 simulations

```{r ten_sims, echo=FALSE, cache =  TRUE}
all_stats$sim <- factor(all_stats$sim)
all_pairstats$sim <- factor(all_pairstats$sim)
subset_sims <- sample(all_stats$sim, 10)
small_all_stats <- filter(all_stats, sim %in% subset_sims)
# Plot values of observed statistics for each simulation replicate
ggplot() +
  geom_point(data = small_all_stats, aes(x = rep, y = het, col = sim)) +
  geom_point(data = observed_stats, aes(x = "observed", y = het)) +
  theme(axis.text.x=element_blank())
small_pairstats <- filter(all_pairstats, sim %in% subset_sims)
ggplot() +
  geom_point(data = small_pairstats, aes(x = rep, y = dxy, col = sim)) +
  geom_point(data = observed_pairstats, aes(x = "observed", y = dxy))+
  theme(axis.text.x=element_blank())
```

## PCA with unnormalized data


```{r pca, echo=TRUE, cache =  TRUE}
sum(is.na(stats)) # should be 0

pc_stats <- prcomp(stats[,-1])
for (k in 1:4) {
  rep_info[[paste0("PC", k)]] <- pc_stats$x[,k]
}

# Convert observed data into PC space
pc_observed <- predict(pc_stats, data.frame(t(observed)))
ggplot(rep_info) +
  geom_point(aes(x = PC1, y = PC2, col = sim)) +
  geom_point(aes(x = pc_observed[,'PC1'], y = pc_observed[,'PC2'])) +
  theme(legend.position="none")
ggplot(rep_info) +
  geom_point(aes(x = PC3, y = PC4, col = sim)) +
  geom_point(aes(x = pc_observed[,'PC3'], y = pc_observed[,'PC4'])) +
  theme(legend.position="none")
```

The loadings on PC1 are all positive and roughly constant

```{r pca_loadings, echo=TRUE, cache =  TRUE}
plot(pc_stats$rotation[,'PC1'])
plot(pc_stats$rotation[,'PC2'])
```

## Look carefully at one SLiM simulation

### Choose a simulation close to observed data

First calculate euclidean distance between all simulated statistics and observed statistics, then choose closest one

```{r choose_close, echo=TRUE, cache =  TRUE}
sim_obs_dist <- rowSums((observed-stats)^2)
min_index <- which.min(sim_obs_dist)
min_info <- rep_info[min_index,]
min_rep <- min_info$rep
min_sim <- min_info$sim
min_recap <- min_info$recap
min_mut <- min_info$mut
min_sample <- min_info$sample

```

### Ten closest recap replicates

```{r close, cache = TRUE}
sim_order <- order(sim_obs_dist)
subset_sims <- unique(rep_info$sim[sim_order[1:10]])
small_all_stats <- filter(all_stats, sim %in% subset_sims)
# Plot values of observed statistics for each simulation replicate
ggplot() +
  geom_point(data = small_all_stats, aes(x = rep, y = het, col = sim)) +
  geom_point(data = observed_stats, aes(x = "observed", y = het)) +
  theme(axis.text.x=element_blank())
small_pairstats <- filter(all_pairstats, sim %in% subset_sims)
ggplot() +
  geom_point(data = small_pairstats, aes(x = rep, y = dxy, col = sim)) +
  geom_point(data = observed_pairstats, aes(x = "observed", y = dxy))+
  theme(axis.text.x=element_blank())
```

### Genetic stats for single closest recapitation replicate

```{r sim_plots, echo=TRUE, cache =  TRUE}
sim_pairstats <- dplyr::filter(all_pairstats, rep == min_rep)
sim_stats <- dplyr::filter(all_stats, rep == min_rep)
ggplot(sim_stats, aes(x = pct, y = het, col = "simulated")) + 
  geom_point() + 
  geom_point(data = observed_stats, aes(x = pct, y = het, col = "observed"))
ggplot(observed_pairstats, aes(x = pair_name, y = dxy, col = "observed")) + 
  geom_point() +
  geom_point(data = sim_pairstats, aes(x = pair_name, y = dxy, col = "simulated"))

```

### Genetic stats for all recapitation replicates

```{r sim_plots2, echo=TRUE}
sim_pairstats <- dplyr::filter(all_pairstats, sim == min_sim)
sim_stats <- dplyr::filter(all_stats, sim == min_sim)
ggplot(sim_stats, aes(x = pct, y = het)) + 
  geom_point() + 
  geom_point(data = observed_stats, aes(x = pct, y = het, col = "observed"))
ggplot(sim_pairstats, aes(x = pair_name, y = dxy, col = "simulated")) + 
  geom_point() + 
  geom_point(data = observed_pairstats, aes(x = pair_name, y = dxy, col = "observed"))
```

### Try running ABC

Before LGM:
CS, AS - migration rates
NE - ancestral population size
Na, Nc, Ns - population sizes
NE bigger than N

Do some posterior predictive checks


```{r abc, echo = FALSE}
library(abc)

#!!!! Why are there NA mutation rates? !!!!
est_params <- c("T2", "T1", "CS", "AS", "NE", "Na", "Nc", "Ns", "mut_rate", "POP_SIZE", "P_D", "DISPERSAL_SIGMA", "YEAR_SHAPE")
na_reps <- which(is.na(rep_info$mut_rate))
abc_obs <- observed
abc_sims <- stats %>% filter(!is.na(rep_info$mut_rate))
abc_rep_info <- dplyr::select(rep_info, est_params) %>% filter(!is.na(mut_rate))
abc_res2 <- abc(abc_obs, abc_rep_info, abc_sims , tol = 0.1, method = "rejection")
# This only works with a higher tolerance, why?
abc_res <- abc(abc_obs, abc_rep_info, abc_sims , tol = 0.2, method = "loclinear")
abc_res3 <- abc(abc_obs, abc_rep_info, abc_sims , tol = 0.1, method = "ridge")
```

```{r plot_abc}
abc_vals <- data.frame(abc_res$unadj.values)
rej_post <- abc_vals %>% pivot_longer(everything(), names_to = "parameter")
adj_post <- abc_vals %>% pivot_longer(everything(), names_to = "parameter")
priors <- rep_info %>% dplyr::select(est_params) %>% pivot_longer(everything(), names_to = "parameter")

ggplot(adj_post, aes(x = value, y = after_stat(density))) +
  geom_histogram() +
  geom_histogram(data = priors, fill = "blue", alpha = 0.5) +
  facet_wrap(~parameter, scales = "free")

ggplot(priors, aes(x = value, y = after_stat(density))) +
  geom_density(fill = "blue", alpha = 0.5) +
  geom_density(data = adj_post) +
  facet_wrap(~parameter, scales = "free")
```

```{r}
pairs(abc_vals)
ggplot(abc_vals, aes(x = POP_SIZE, y = P_D)) +
  geom_point()
ggplot(abc_vals, aes(x = POP_SIZE, y = DISPERSAL_SIGMA)) +
  geom_point()
```

```{r}
estimates <- apply(abc_res$adj.values, 2, median)
```

### Test on held out simulation

```{r test_abc}
# Hold out a simulation and do ABC
# Simulation we're using as observation
set.seed(100)
(sim_observed_num <- sample(levels(rep_info$sim), 1))
(sim_observed_rows <- which(rep_info$sim == sim_observed_num))
(sim_observed <- stats[sim_observed_rows,])
```